---
title: "R programming coursework"
author: "2228202"
date: "2022-11-28"
output: rmarkdown::github_document
---

# Install

```{r install, eval=FALSE}
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("rstatix")
install.packages("ggpubr")
install.packages("knitr")
install.packages("performance")
install.packages("interactions")
install.packages("jtools")
install.packages("wesanderson")
install.packages("vroom")
install.packages("gridExtra")
install.packages("ggh4x")
install.packages('fitdistrplus')
install.packages("glmmTMB")
install.packages("MuMIn")
install.packages("DHARMa")


```

## Load

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("tidyverse")
library("ggplot2")
library("rstatix")
library("ggpubr")
library("knitr")
library("performance")
library("interactions")
library("jtools")
library("wesanderson")
library("vroom")
library("gridExtra")
library("ggh4x")
library('fitdistrplus')
library("glmmTMB")
library("MuMIn")
library("DHARMa")
```

# Introduction 


Investigate if summer NAO (sNAO) has an effect on body mass of *Ovis aries* and determine other factors.

## Hypothesis 

* Null - The sNAO has no effect on body mass within the population.
* Alternative - NAO and sNAO determines weight within the population.


# Exploratory Data Analysis

## Import


```{r import}

#Vroom package to load data.

#sets wd to location of rmd file R studio only not console
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))


# load in zipped gz files
mass_data <- vroom("../Data/mass_fh22528.tsv.gz")
nao_data <- vroom("../Data/NAO_fh22528.tsv.gz")
```


## Data Structure


```{r data head and tail}
#kable function presents in a table.
knitr::kable(head(mass_data))

knitr::kable(head(nao_data))

```

```{r data structure}

# view the data types 
dplyr::glimpse(mass_data)

dplyr::glimpse(nao_data)

```


### mass_data 

* response variable - weight (kgs) - data has been normalised.

* predictor variables
  + Sex 
  + NAO 
  + Age 

### nao_data
* Response variable - NAO
  + negative (low) and positive (high).
* predictor variable
  + year
  + month

## Visualisations


```{r EDA visuals}

# Group NAO by year 
grouped_NAO <- nao_data %>% 
  group_by(year) %>% 
  summarise(mean_NAO = mean(NAO))

ggplot(grouped_NAO, aes(y = mean_NAO, x = year )) +
  geom_line(col = "lightgrey") +
  geom_point(col = "darkorchid3") +
  theme_classic() +
  ylab("difference in mean NAO (hPa)") +
  ggtitle("Yearly NAO difference")
  

```
**Figure 1.** Yearly mean NAO difference -
  The figure, indicates that there is a clear positive relationship between NAO and year.There is a high level of fluctuation indicated by    the peaks and troughs.


```{R summer check}

ggplot(nao_data, aes(y = NAO, x = year, col = month)) +
  geom_point() +
  theme_classic() +
  ggtitle("Yearly NAO difference by month")




```

**Figure 2.** Yearly NAO difference by month
There is a clear pattern of NAO pressure being much more variable in June, July and august.



```{r EDA jitter}

#analysis of NAO data points by month

#converts month character so it can be ordered.
nao_data$month = factor(nao_data$month,levels = month.name)

# NAO difference by month - boxplot and   jitter
ggplot(nao_data, aes(y = NAO, 
                     x = factor(nao_data$month,levels = month.name), fill = month)) +
  geom_boxplot(alpha = 0.3, outlier.shape=NA) +
  geom_point(position=position_jitterdodge()) +
  theme_classic() +
  ylab("difference NAO (hPa)") +
  xlab("Month") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        ) +
  ggtitle("Monthly NAO difference")
  
  

```

**Figure 3.** Monthly NAO difference  -
This shows that NAO has much more variation within the months of June-August.


## Mass data analysis

### Check NAs


```{r missing values, warning=FALSE}

#check for NA values and present results in a table

knitr::kable(mass_data %>%
  dplyr::select(everything()) %>%  #select all columns
  summarise_all(funs(sum(is.na(.)),n = n())) %>% #apply to all columns
  dplyr::select(c(1:9))) # only show 1 sample total

knitr::kable(nao_data %>%
  dplyr::select(everything()) %>%  #select all columns
  summarise_all(funs(sum(is.na(.)),n = n())) %>% #apply to all columns
   dplyr::select(c(1:4)))# only show 1 sample total

```


```{r}
#remove_NAs
mass_data_clean <- mass_data %>% 
  dplyr::filter(!is.na(weight))  #remove NAs

```


```{r weight histogram}
# check for normal distribution of data of our response variable
ggplot(mass_data_clean, aes(x = weight, alpha=after_stat(count))) + 
  geom_histogram(fill = "darkolivegreen4", binwidth = 1) +
  theme_classic() +
  theme(legend.position = "none")

```

**Figure 4.** Distribution of weight difference -
This figure shows how weight difference has a central tendency towards 1kg.

```{r weight qqplot, warning=FALSE}

ggqqplot(mass_data_clean, x = "weight", color = "dodgerblue4") + theme_classic()

```

**Figure 5.** Weight qqplot  -
At the extreme ends weight data does not fit very well to a normal distribution, it is lightly tailed.




**Figure 6.** Weight difference by age and sex  -
Shows moderate positive correlations between age and weight difference for both sex groups.

```{r}

#weight and sex
ggplot(mass_data_clean, aes(x= sex, y = weight, fill = sex)) + 
  geom_boxplot() + theme_classic() +
  scale_fill_manual(values = wes_palette("GrandBudapest1"))+
  scale_color_manual(values = wes_palette("GrandBudapest1"))

```

**Figure 7. ** Sex and weight difference  -
The difference in mean weight between the male and female groups.

```{r}
#weight, n_babies, facet by sex
ggplot(mass_data_clean, aes(x = as.factor(n_babies), y = weight, fill = as.factor(n_babies))) + 
  geom_violin() +
  geom_boxplot(width=0.2, fill = "white") +
  facet_grid(~ sex) +
  theme_classic() +
  scale_fill_manual(values = wes_palette("Zissou1"))+
  scale_color_manual(values = wes_palette("Zissou1")) +
  xlab(" Number of babies") +
  stat_summary( fun = mean, geom="point") +
  theme(legend.position="none")

```

**Figure 8. ** Number of babies and weight difference  -
The number of babies does not influence a difference in weight.


``` {R weight, year and age}

#weight, age and year
ggplot(mass_data_clean, aes(x= year, y = weight, col = age)) + 
  geom_point() + theme_classic()
  #scale_fill_manual(values = wes_palette("GrandBudapest1"))+
  #scale_color_manual(values = wes_palette("GrandBudapest1"))


```
Figure 9 - Weight by Year and Age
Older sheep do have larger weights.


* **Summary** - Year, age and sex are interesting variables





# Data transformation

The mass data needs to be joined to the NAO data by the year column.



```{r}


# Join yearly mass to yearly NAO data.
mass_NAO_df <- dplyr::left_join(mass_data_clean, nao_data,
                                     by = c("year"= "year"))
#view the transformation
knitr::kable(head(mass_NAO_df))

```
 

```{R transform to sNAO category}

#### lets try to create a summer categorical variable to group by ###
## 28/12/2022

#filter for the summer months with increased NAO variation
summer_filter <- c("June", "July", "August")

### creating a category of sNAO and not sNAO for visualisaton
mass_NAO_df <- mass_NAO_df %>%
  mutate(nao_type = case_when(month %in% summer_filter ~ "summer_NAO", #in summer
                              !month %in% summer_filter ~ "not_summer_NAO"))#not in summer

knitr::kable(head(mass_NAO_df))

```



## Data Visualisation
Data visualation after the join.

```{r point mean weight & mean NAO}
#point data
ggplot(data = mass_NAO_df, 
       aes(x = NAO , y = weight, colour = nao_type )) + 
  geom_point() + theme_classic()

```

**Figure 10 ** weight by summer NAO
A view of the data after joining Mass and NAO data via year show clustering of non-summer months.



# Generalised linear models (GLMs)

Examine casual relationships using GLMs.

* Variables of interest.
  + sex
  + NAO
  + year
  + age
  + month
  


## Deciding distributions


```{r fitdist }

#check oberserved weigth differences, against the normal distribution.
plot(fitdist(mass_NAO_df$weight, "norm"))


#summary statistics, skewness is low.
descdist(mass_NAO_df$weight, boot = 1000)

```

**Figure 11.** Fitdist the normal distribution -
The normal distribution can be used in the model,there are negative values which limit other families.



## Creating Models 

```{R GLM creating models }
# weight - predictors - NAO
mod1 <- glm(weight ~ NAO,
            data = mass_NAO_df,
            family = "gaussian")

# weight - predictors - NAO + sex 
mod2 <-  glm(weight ~ NAO + sex,
            data = mass_NAO_df,
            family = "gaussian")

# weight - predictors - NAO + sex 
mod3 <-  glm(weight ~ NAO + sex * age,
            data = mass_NAO_df,
            family = "gaussian") 


# weight - predictors - NAO + sex 
mod4 <-  glm(weight ~ NAO + sex + age * year,
            data = mass_NAO_df,
            family = "gaussian")


# null model for comparison
null_mod <- glm(weight ~ 1,
            data = mass_NAO_df,
            family = "gaussian")

        
```

## Assessing Model fit

```{R Comparing models}

# The performance package can compare models
compare_performance(mod1, mod2, mod3, mod4) %>% kable()

compare_performance(mod1, mod2, mod3, mod4, rank = TRUE) %>% kable()

```


**Table 1.** - A summary of GLM model formulas and ranking.
Model 3 and 4 are the best performing models.



```{r visual model performance}

plot(compare_performance(mod1, mod2, mod3, mod4))

```

**Figure 12.** Model Ranking -
A visual representation of model ranking, while a clear lowest AIC value has been determined in model 3, model 6 shows promising interactions of nao type variable and sex.


```{r heteroscedasticity }
#Checking for heteroscedasticity
check_heteroscedasticity(mod1)
check_heteroscedasticity(mod2)
check_heteroscedasticity(mod3)
check_heteroscedasticity(mod4)

```


## Model output

```{r model outputs 4}

summary(mod3)
summary(mod4)

```


Model 3 and 4 do have significant p-values for sex and age predictor variables.


```{r annova as model }

# comparing deviance of models by order of ranking.

anova(mod1, mod2, mod3, test = "Chisq")

```





```{r null model }

# lets also compare predictors against a null model
anova(mod4, null_mod, test= "Chisq")
```





```{r model 4.1}
# computes a statistically significant p-value for chi-squared and formats the output better.
summ(mod4, confint = TRUE)

```

```{r}

export_summs(mod4)

```



## Predicted vs observed


```{r}


#Lets compare predicted vs observed values from some promising models.


#model 3
##visualize the fitted vs observed values for model 3 
fit_data4 <- data.frame("predicted"= mod4$fitted.values,
                       "observed" = mass_NAO_df$weight)
##plot them
po1 <- ggplot(fit_data4, aes(x=observed, y=predicted)) + 
  geom_point(col = "lightgreen") + 
  ##add a 1:1 line
  geom_abline(intercept = 0) +
  ##add a linear regression to your data
  geom_smooth(method="lm",
              col="lightblue",
              se = F) +
  #colour adjustment
  ##set the theme
  theme_classic() +
  theme(legend.position="none") +
  ggtitle("a) Model 3")



#model 3.2 without interactions
##visualize the fitted vs observed values for model 3 
fit_data_3 <- data.frame("predicted"= mod3$fitted.values,
                       "observed" = mass_NAO_df$weight)
##plot them
po2 <- ggplot(fit_data_3, aes(x=observed, y=predicted)) + 
  geom_point(col = "firebrick") + 
  ##add a 1:1 line
  geom_abline(intercept = 0) +
  ##add a linear regression to your data
  geom_smooth(method="lm",
              col="lightblue",
              se = F) +
  #colour adjustment
  ##set the theme
  theme_classic() +
  theme(legend.position="none" ) +
  ggtitle("b) Model 4") +
  theme(axis.title.y =element_blank()) # remove dup axis




grid.arrange(po1,po2, nrow = 1)



```


**Figure 13.** Predicted vs observed weight values -


## visualising the models

The check models package can be used to visualise model performance.

```{r check model 3}

check_model(mod3)

```

**Figure 14.** Performance package visualisation of model 3 -
This visualization shows that the observed data fit relatively well to the model prediction.
The error residuals fit a normal distribution via the qqplot but does tail off at the extremes.
However, the model is presenting heteroscedasticity which is due to large variation in the dependent variable.


```{r check model 4}

check_model(mod4)

```


# Discussion

To conclude, a positive relationship between weight difference and mean NAO has been identified (Figure 9). Sex and age are predictors of weight with males presenting a decrease in response (Figure 7). 

There is evidence to reject the null hypothesis and claim that NAO is a variable that is linked to weight difference in *Oves ares*. The best GLM models identified in GLM analysis were Model 3 and 4. Further analysis using GLMMs is required to examine random effects of year and age.













 
 
 
 