---
title: "R programming coursework"
author: "2228202"
date: "2022-11-28"
output: rmarkdown::github_document
---

# Install

```{r install, eval=FALSE}
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("rstatix")
install.packages("ggpubr")
install.packages("knitr")
install.packages("performance")
install.packages("interactions")
install.packages("jtools")
install.packages("wesanderson")
install.packages("vroom")
install.packages("gridExtra")
install.packages("ggh4x")
install.packages('fitdistrplus')
install.packages("glmmTMB")
install.packages("MuMIn")
install.packages("DHARMa")


```

## Load

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("tidyverse")
library("ggplot2")
library("rstatix")
library("ggpubr")
library("knitr")
library("performance")
library("interactions")
library("jtools")
library("wesanderson")
library("vroom")
library("gridExtra")
library("ggh4x")
library('fitdistrplus')
library("glmmTMB")
library("MuMIn")
library("DHARMa")
```

# Introduction 


Investigate if summer NAO (sNAO) has an effect on body mass of *Ovis aries* and determine other factors.

## Hypothesis 

* Null - The sNAO has no effect on body mass within the population.
* Alternative - NAO and sNAO determines weight within the population.


# Exploratory Data Analysis

## Import


```{r import}

#Vroom package to load data.

#sets wd to location of rmd file R studio only not console
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))


# load in zipped gz files
mass_data <- vroom("../Data/mass_fh22528.tsv.gz")
nao_data <- vroom("../Data/NAO_fh22528.tsv.gz")
```


## Data Structure


```{r data head and tail}
#kable function presents in a table.
knitr::kable(head(mass_data))

knitr::kable(head(nao_data))

```

```{r data structure}

# view the data types 
dplyr::glimpse(mass_data)

dplyr::glimpse(nao_data)

```


### mass_data 

* response variable - weight (kgs) - data has been normalised.

* predictor variables
  + Sex 
  + NAO 
  + Age 

### nao_data
* Response variable - NAO
  + negative (low) and positive (high).
* predictor variable
  + year
  + month

## Visualisations


```{r EDA visuals}

# Group NAO by year 
grouped_NAO <- nao_data %>% 
  group_by(year) %>% 
  summarise(mean_NAO = mean(NAO))

ggplot(grouped_NAO, aes(y = mean_NAO, x = year )) +
  geom_line(col = "lightgrey") +
  geom_point(col = "darkorchid3") +
  theme_classic() +
  ylab("difference in mean NAO (hPa)") +
  ggtitle("Yearly NAO difference")
  

```
**Figure 1.** Yearly mean NAO difference -
  The figure, indicates that there is a clear positive relationship between NAO and year.There is a high level of fluctuation indicated by    the peaks and troughs.


```{R summer check}

ggplot(nao_data, aes(y = NAO, x = year, col = month)) +
  geom_point() +
  theme_classic() +
  ggtitle("Yearly NAO difference by month")




```

**Figure 2.** Yearly NAO difference by month
There is a clear pattern of NAO pressure being much more variable in June, July and august.



```{r EDA jitter}

#analysis of NAO data points by month

#converts month character so it can be ordered.
nao_data$month = factor(nao_data$month,levels = month.name)

# NAO difference by month - boxplot and   jitter
ggplot(nao_data, aes(y = NAO, 
                     x = factor(nao_data$month,levels = month.name), fill = month)) +
  geom_boxplot(alpha = 0.3, outlier.shape=NA) +
  geom_point(position=position_jitterdodge()) +
  theme_classic() +
  ylab("difference NAO (hPa)") +
  xlab("Month") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        ) +
  ggtitle("Monthly NAO difference")
  
  

```

**Figure 2.** Monthly NAO difference  -
This shows that NAO has much more variation within the months of June-August.


## Mass data analysis

### Check NAs


```{r missing values, warning=FALSE}

#check for NA values and present results in a table

knitr::kable(mass_data %>%
  dplyr::select(everything()) %>%  #select all columns
  summarise_all(funs(sum(is.na(.)),n = n())) %>% #apply to all columns
  dplyr::select(c(1:9))) # only show 1 sample total

knitr::kable(nao_data %>%
  dplyr::select(everything()) %>%  #select all columns
  summarise_all(funs(sum(is.na(.)),n = n())) %>% #apply to all columns
   dplyr::select(c(1:4)))# only show 1 sample total

```


```{r}
#remove_NAs
mass_data_clean <- mass_data %>% 
  dplyr::filter(!is.na(weight))  #remove NAs

```


```{r weight histogram}
# check for normal distribution of data of our response variable
ggplot(mass_data_clean, aes(x = weight, alpha=after_stat(count))) + 
  geom_histogram(fill = "darkolivegreen4", binwidth = 1) +
  theme_classic() +
  theme(legend.position = "none")

```

**Figure 3.** Distribution of weight difference -
This figure shows how weight difference has a central tendency towards 1kg.

```{r weight qqplot, warning=FALSE}

ggqqplot(mass_data_clean, x = "weight", color = "dodgerblue4") + theme_classic()

```

**Figure 4.** Weight qqplot  -
At the extreme ends weight data does not fit very well to a normal distribution, it is lightly tailed.




**Figure 5.** Weight difference by age and sex  -
Shows moderate positive correlations between age and weight difference for both sex groups.

```{r}

#weight and sex
ggplot(mass_data_clean, aes(x= sex, y = weight, fill = sex)) + 
  geom_boxplot() + theme_classic() +
  scale_fill_manual(values = wes_palette("GrandBudapest1"))+
  scale_color_manual(values = wes_palette("GrandBudapest1"))

```

**Figure 7. ** Sex and weight difference  -
The difference in mean weight between the male and female groups.

```{r}
#weight, n_babies, facet by sex
ggplot(mass_data_clean, aes(x = as.factor(n_babies), y = weight, fill = as.factor(n_babies))) + 
  geom_violin() +
  geom_boxplot(width=0.2, fill = "white") +
  facet_grid(~ sex) +
  theme_classic() +
  scale_fill_manual(values = wes_palette("Zissou1"))+
  scale_color_manual(values = wes_palette("Zissou1")) +
  xlab(" Number of babies") +
  stat_summary( fun = mean, geom="point") +
  theme(legend.position="none")

```

**Figure 8. ** Number of babies and weight difference  -
The number of babies does not influence a difference in weight.


``` {R weight, year and age}

#weight, age and year
ggplot(mass_data_clean, aes(x= year, y = weight, col = age)) + 
  geom_point() + theme_classic()
  #scale_fill_manual(values = wes_palette("GrandBudapest1"))+
  #scale_color_manual(values = wes_palette("GrandBudapest1"))


```
Figure 9 - Weight by Year and Age
Older sheep do have larger weights.


* **Summary** - Year, age and sex are interesting variables





# Data transformation

The mass data needs to be joined to the NAO data by the year column.



```{r}


# Join yearly mass to yearly NAO data.
mass_NAO_df <- dplyr::left_join(mass_data_clean, nao_data,
                                     by = c("year"= "year"))
#view the transformation
knitr::kable(head(mass_NAO_df))

```
 

```{R transform to sNAO category}

#### lets try to create a summer categorical variable to group by ###
## 28/12/2022

#filter for the summer months with increased NAO variation
summer_filter <- c("June", "July", "August")

### creating a category of sNAO and not sNAO for visualisaton
mass_NAO_df <- mass_NAO_df %>%
  mutate(nao_type = case_when(month %in% summer_filter ~ "summer_NAO", #in summer
                              !month %in% summer_filter ~ "not_summer_NAO"))#not in summer

knitr::kable(head(mass_NAO_df))

```



## Data Visualisation
Data visualation after the join.

```{r point mean weight & mean NAO}
#point data
ggplot(data = mass_NAO_df, 
       aes(x = NAO , y = weight, colour = nao_type )) + 
  geom_point() + theme_classic()
   #scale_fill_manual(values = wes_palette("Royal1"))+
  #scale_color_manual(values = wes_palette("Royal1"))
  
```

**Figure 10 ** weight by summer NAO
A view of the data after joining Mass and NAO data via year show clustering of non-summer months.



# Generalised linear models (GLMs)

Examine casual relationships using GLMs.

* Variables of interest.
  + sex
  + NAO
  + year
  + age
  + month
  


## Deciding distributions


```{r fitdist }

#check oberserved weigth differences, against the normal distribution.
plot(fitdist(mass_NAO_df$weight, "norm"))


#summary statistics, skewness is low.
descdist(mass_NAO_df$weight, boot = 1000)

```

**Figure 15.** Fitdist the normal distribution -
The normal distribution can be used in the model,there are negative values which limit other families.



## Creating Models 

```{R GLM creating models }
# weight - predictors - NAO
mod1 <- glm(weight ~ NAO,
            data = mass_NAO_df,
            family = "gaussian")

# weight - predictors - NAO + sex 
mod2 <-  glm(weight ~ NAO + sex,
            data = mass_NAO_df,
            family = "gaussian")

# weight - predictors - NAO + sex 
mod3 <-  glm(weight ~ NAO + sex * age,
            data = mass_NAO_df,
            family = "gaussian") 


# weight - predictors - NAO + sex 
mod4 <-  glm(weight ~ NAO + sex + age * year,
            data = mass_NAO_df,
            family = "gaussian")


# null model for comparison
null_mod <- glm(weight ~ 1,
            data = mass_NAO_df,
            family = "gaussian")

        
```

## Assessing Model fit

```{R Comparing models}

# The performance package can compare models
compare_performance(mod1, mod2, mod3, mod4) %>% kable()

compare_performance(mod1, mod2, mod3, mod4, rank = TRUE) %>% kable()

```


**Table 1.** - A summary of GLM model formulas and ranking.
Model 3 and 4 are the best performing models.



```{r visual model performance}

plot(compare_performance(mod1, mod2, mod3, mod4))

```

**Figure 16.** Model Ranking -
A visual representation of model ranking, while a clear lowest AIC value has been determined in model 3, model 6 shows promising interactions of nao type variable and sex.


```{r heteroscedasticity }
#Checking for heteroscedasticity
check_heteroscedasticity(mod1)
check_heteroscedasticity(mod2)
check_heteroscedasticity(mod3)
check_heteroscedasticity(mod4)

```


## Model output

```{r model outputs 4}

summary(mod3)
summary(mod4)

```
* **coefficients**

The sex group male has a significant difference because the p-value is < 0.05.
The male group sees an average logit decrease of -13.1 in the response variable weight. It is interesting that the equivalent in females observes an increase of 4.13. 

The model provides evidence that male weight is more negatively affected by NAO.

The age_yrs variable has a p-value < 0.05, which suggests age affects the weight response variable. 
The interaction between sex and age is not significant.


* **Null and residual deviance**

There is a large decrease in residual deviance compared to the null deviance suggesting the predictors have influence on weight.
A limitation is the residual deviance is high when compared to the residual degrees of freedom.


```{r annova as model }

# comparing deviance of models by order of ranking.

anova(mod1, mod2, mod3, test = "Chisq")

```





```{r null model }

# lets also compare predictors against a null model
anova(mod4, null_mod, test= "Chisq")
```





```{r model 4.1}
# computes a statistically significant p-value for chi-squared and formats the output better.
summ(mod4, confint = TRUE)

```

```{r}

export_summs(mod4)

```



## Predicted vs observed


```{r}


#Lets compare predicted vs observed values from some promising models.


#model 3
##visualize the fitted vs observed values for model 3 
fit_data4 <- data.frame("predicted"= mod4$fitted.values,
                       "observed" = mass_NAO_df$weight)
##plot them
po1 <- ggplot(fit_data4, aes(x=observed, y=predicted)) + 
  geom_point(col = "lightgreen") + 
  ##add a 1:1 line
  geom_abline(intercept = 0) +
  ##add a linear regression to your data
  geom_smooth(method="lm",
              col="lightblue",
              se = F) +
  #colour adjustment
  ##set the theme
  theme_classic() +
  theme(legend.position="none") +
  ggtitle("a) Model 3")



#model 3.2 without interactions
##visualize the fitted vs observed values for model 3 
fit_data_3 <- data.frame("predicted"= mod3$fitted.values,
                       "observed" = mass_NAO_df$weight)
##plot them
po2 <- ggplot(fit_data_3, aes(x=observed, y=predicted)) + 
  geom_point(col = "firebrick") + 
  ##add a 1:1 line
  geom_abline(intercept = 0) +
  ##add a linear regression to your data
  geom_smooth(method="lm",
              col="lightblue",
              se = F) +
  #colour adjustment
  ##set the theme
  theme_classic() +
  theme(legend.position="none" ) +
  ggtitle("b) Model 3.2") +
  theme(axis.title.y =element_blank()) # remove dup axis




grid.arrange(po1,po2, nrow = 1)



```


**Figure 17.** Predicted vs observed weight values -


## visualising the models

The check models package can be used to visualise model performance.

```{r check model}

check_model(mod3)

```

**Figure 18.** Performance package visualisation of model 3 -
This visualization shows that the observed data fit relatively well to the model prediction.
The error residuals fit a normal distribution via the qqplot.
However, the model is presenting heteroscedasticity which is due to large variation in the dependent variable.


```{r check model}

check_model(mod4)

```






## Interactions



```{r interact plot model 6 }


#plot variation interaction of year and mean NAO on weight change
interact_plot(mod4,
              pred = age,
              modx = year,
              plot.points = TRUE,
             # point.shape = TRUE, # different shape for differennt levels of the moderator
             # partial.residuals = TRUE,
              jitter = 0.1) # if two data points are on top one another, this moves them apart by little)


```

**Figure 20.2** - Investigating the interactions between NAO and year.




```{r slope analysis }

summary(mod4)

#slope analysis of year.
  ss <- sim_slopes(mod4,
           pred = sex,
           modx = year,
           modx.values = c(1990, 2000,2010,2020),
           johnson_neyman = FALSE,
           )

ss
plot(ss)

```











# Generalised linear mixed models (GLMMs)


* Random effects (potential)
  + site
  + number of babies
  + age/ year






## Fitting GLMMs -
```{r GLMM}

#### Random intercepts  #####


#site as a random effect
m_mod1 <- glmmTMB(mean_weight_diff_kg ~ mean_nao + sex * age_yrs +
                    (1|site), #random intercept effects
                  data=grp_mass_pred_NAO_GLMM,
                  family="gaussian")



# number of babies as a random effect
m_mod2 <-  glmmTMB(mean_weight_diff_kg ~ mean_nao + sex * age_yrs +
                   (1|n_babies), #random intercept effects
                   data=grp_mass_pred_NAO_GLMM,
                   family="gaussian")

# age as a random effect
m_mod3 <- glmmTMB(mean_weight_diff_kg ~ mean_nao + sex +
                    (1|age_yrs ), #random intercept effects
                  data=grp_mass_pred_NAO_GLMM,
                  family="gaussian")

# year as a random effect
m_mod5 <- glmmTMB(mean_weight_diff_kg ~ mean_nao + sex * age_yrs +
                    (1|year), #random intercept effects
                  data=grp_mass_pred_NAO_GLMM,
                  family="gaussian")

# year as a random effect with no age predictor
m_mod5.1 <- glmmTMB(mean_weight_diff_kg ~ mean_nao + sex +
                    (1|year), #random intercept effects
                  data=grp_mass_pred_NAO_GLMM,
                  family="gaussian")

```


## GLMM outputs


```{r site and no numbers glmm, warning=FALSE}

# there is very little difference between the marginal and conditional values for these two predictors.

a <-  MuMIn::r.squaredGLMM(m_mod1)
b <-  MuMIn::r.squaredGLMM(m_mod2)

rbind(a,b) %>% kable()

```

Site and number of babies are not a random effect.

```{r, r2 GLM model 5 comp, warning=FALSE}

c <-  MuMIn::r.squaredGLMM(m_mod5)
d <- MuMIn::r.squaredGLMM(m_mod5.1)

rbind(c,d) %>% kable()

```

Year has the highest variance suggesting a random effect on weight change.

## Nesting 


```{r}

# year as a random effect nested with age.
m_mod5.2 <- glmmTMB(mean_weight_diff_kg ~ mean_nao + sex +
                    (1|year/age_yrs), #random intercept effects
                  data=grp_mass_pred_NAO_GLMM,
                  family="gaussian")


MuMIn::r.squaredGLMM(m_mod5.2)

```

```{r model comparison} 

# this nicely compares models glmm model results.
(parcomp <- cbind(m_mod2=unlist(fixef(m_mod2)),
                  m_mod5.1=unlist(fixef(m_mod5.1)),
                  m_mod5.2=unlist(fixef(m_mod5.2)))) %>% kable()

```



```{r}


anova(m_mod1, m_mod5.1, m_mod5.2)
      

```


```{r GLMM outputs }

summary(m_mod5.2)

```


Year nested with age has a more varied maginal and conditional.

```{r}

testDispersion(m_mod5.1) 
testDispersion(m_mod5.2) 

```


## Plotting model
```{r}

m_mod5.1_sim <- simulateResiduals(m_mod5.1, n = 1000)
m_mod5.2_sim <- simulateResiduals(m_mod5.2, n = 1000)

plot(m_mod5.1_sim)
plot(m_mod5.2_sim) 


```

**Figure 21.** - The age nesting causes a significant dispersion result which could indicate a mis-specified model.


### coefficients


```{r}

## reshape the outputs of the model into a tidy format:
tidy_mod5.1 <- broom.mixed::tidy(m_mod5.1, conf.int = TRUE) %>% 
  ##clean up the random effect names
  mutate(term=ifelse(grepl("sd__(Int",term,
                           fixed=TRUE),
                     paste(group,term,sep="."),
                     term))


##filter and sort the data
sorted_tidy_mod5.1 <- 
  tidy_mod5.1 %>%
  filter(term!="(Intercept)") %>%
  mutate(term=factor(term, levels = rev(term)))

##plot it out
ggplot(sorted_tidy_mod5.1, aes(x=estimate, y=term)) + 
  geom_errorbar(aes(xmin=conf.low, xmax=conf.high),                       
                width=0.3, 
                col="grey50",
                alpha=0.7) +
  geom_point(col="coral", size=5, alpha=0.9) + 
  geom_vline(xintercept=0, lty="dashed", alpha=0.3) +
  theme_bw() +
  ylab("") +
  xlab("Coefficient estimate")


```
**Figure 22.** Visualisations for the model coefficients.

## Yearly temporal correlations


```{r temporal auto correlation}
# Individuals have been sampled over a long period which can lead to artificial correlations.
# temporal correlations model - year

m_mod4.1 <- glmmTMB(mean_weight_diff_kg ~ mean_NAO + 
                      (1|year), #random intercept effects)
                    data=grouped_mass_NAO, # time must be unique so predictors removed.
                  family="gaussian")

summary(m_mod4.1)

m_mod4.1_sim <- simulateResiduals(m_mod4.1, n = 1000)


testTemporalAutocorrelation(m_mod4.1_sim,
                            time = grouped_mass_NAO$year,
                            plot = TRUE)

```

**Figure 23.** - temporal auto correlation analysis.
Fitting a simplified model with weight against mean NAO shows fluctuations in residuals overtime, with a p-value of < 0.05.
This analysis has used year as a link between mass and NAO and the most recent years are showing a rise.


## Predicted vs observed.

```{r}



#create df for model 5.1 observed vs predicted
fit_glmm_5.1 <- data.frame("predicted"= predict(m_mod5.1,data = grp_mass_pred_NAO_GLMM, type = "response"),
                        "observed" = grp_mass_pred_NAO_GLMM$mean_weight_diff_kg)


# create df for model 5.2 observed vs predicted
fit_glmm_5.2 <- data.frame("predicted"= predict(m_mod5.2,data = grp_mass_pred_NAO_GLMM, type = "response"),
                       "observed" = grp_mass_pred_NAO_GLMM$mean_weight_diff_kg)


# create df for model 5.2 observed vs predicted
fit_glmm_4.1 <- data.frame("predicted"= predict(m_mod4.1,data = grouped_mass_NAO, type = "response"),
                       "observed" = grouped_mass_NAO$mean_weight_diff_kg)


## Model 5.1 plot the predicted against the observed
glmm_mod5.1 <- ggplot(fit_glmm_5.1, aes(x = observed,
                       y = predicted)) +
  geom_point(col="lightblue") +
  geom_abline(slope = 1) +
  theme_minimal() +
  xlab("Observed") +
  ylab("Predicted") +
  ggtitle("mod5.1")



##Model 5.2 plot the predicted against the observed
glmm_mod5.2 <- ggplot(fit_glmm_5.2, aes(x = observed,
                       y = predicted)) +
  geom_point(col="grey") +
  geom_abline(slope = 1) +
  theme_minimal() +
  xlab("Observed") +
  ylab("Predicted") +
   ggtitle("mod5.2") +
  theme(axis.title.y =element_blank()) # remove dup axis



##Model 4.1 plot the predicted against the observed
glmm_mod4.1 <- ggplot(fit_glmm_4.1, aes(x = observed,
                       y = predicted)) +
  geom_point(col="lightgreen") +
  geom_abline(slope = 1) +
  theme_minimal() +
  xlab("Observed") +
  ylab("Predicted") +
   ggtitle("mod4.1") +
  theme(axis.title.y =element_blank()) # remove dup axis



grid.arrange(glmm_mod5.1, glmm_mod5.2, glmm_mod4.1, nrow = 1)

```

**Figure 18** Predicted vs observed GLMMs.

# Discussion

To conclude, a positive relationship between weight difference and mean NAO has been identified (Figure 9). Sex and age are predictors of weight with males presenting a decrease in response (Figure 7). Figure 13 shows how this relationship correlates higher within the sNAO period.

There is evidence to reject the null hypothesis and claim that NAO is a variable that is linked to weight difference in *Oves ares*.

However, a single causal relationship between sNAO and weight change is not appropriate. 

Year has been estimated as a random effect on weight, but a significant temporal auto correlation is overestimating the relationships. (Figure 23) 

Further investigations could include weight transformations (cube roots) to remove negative values and apply different models or adding an age range category.












 
 
 
 